\pagenumbering{roman}
\thispagestyle{empty}
\maketitle

\newpage
\
\thispagestyle{empty}
\newpage

\thispagestyle{empty}
\begin{abstract}
	With the growing scale of big data, fitting novel statistical models on larger-than-memory datasets becomes correspondingly challenging.
	A variety of techniques exist in an attempt to solve this problem in a general case, with generally subpar results in practice among statisticians.
	In the absence of a sound interface and conceptual structure for designing statistical algorithms for large scale data, the large-scale statistical modelling field has lacked the accessibility and results that it otherwise may have possessed.
	This thesis outlines the development and use of a platform created specifically for the development of statistical models for big datasets.
	Requirements for such a platform are constructed and justified, including a formally specified interface.
	The delivered platform satisfies such requirements, and serves as a functioning example of what form a large scale statistical modelling platform backed by sound conceptual requirements may take.
\end{abstract}
\addcontentsline{toc}{chapter}{Abstract}

\chapter*{\centering \large Acknowledgements}
\thispagestyle{empty}
Thank you --
\\
To my supervisors, Simon Urbanek and Paul Murrell.
I undertook this study to learn under a master, as an apprentice.
Your years of experience shaped my worldview, and will continue to guide me as I go.
Simon, you went far beyond what any supervisor is expected to give, and became a closest friend.
\\
To the family around me -- my wife Caroline, and our children Chloe and James.
This family was born during these student years, and pulled me into adulthood.
You showed me that the future is bright and worth reaching for.
I hope to pull us toward it in turn.
\\
To the family who raised me.
Your education and encouragement made this possible.
Ian -- thank you for keeping me grounded.
Wayne -- your persistent care and reliability stood out when it mattered most.
\\
To my university colleagues -- Adam, Godrick, Emeka -- and most notably, the emeritus Valentine Ibeka.
\\
Valentine -- you made a sociologist out of me, whether I wanted to be one or not.
I only hope I have shaped you in less sinister ways, though I suspect the effect was mutual.
\\
To my friends. This thesis contained many lifetimes; your presence kept me in the right one.
\\
Jonty and Kayla -- the mothparents -- you made it fun. Let's always keep that fun.
\\
Josh, Reuben, Andrew, Aaron, Sean -- you kept me going, each in your own strange way.
\\
To the team at Air New Zealand. Thank you for taking the risk on a doctoral student. I hope it paid off.
\\
This work was supported by the Australian Research Data Commons (ARDC) through the Nectar Research Cloud\cite{nectar2021}.
\addcontentsline{toc}{chapter}{Acknowledgements}

\newpage
\
\thispagestyle{empty}
\newpage

\newpage
\settocdepth{subsection}
\tableofcontents
\newpage \ \newpage
\listoffigures
\newpage
\listoftables
\newpage \ \newpage
\listoflistings
\addcontentsline{toc}{chapter}{List of Listings}

